--- 
author: "Rebecca Barter" 
categories: [R, superheat] 
date: 2017-02-05T09:44:30-08:00 
type: "post" 
description: "First version of superheat now available on CRAN." 
featured: "" 
featuredalt: "" 
featuredpath: "" 
linktitle: "" 
title: "superheat 0.1.0" 
---


superheat 0.1.0 is now available on CRAN. Superheat makes it easy to create extendable, cutomizable, and most importantly, beautiful heatmaps. It has increased flexibility and user-friendliness when compared to alternatives such as `heatmap()` and `pheatmap()`.

For usage options see the [**vignette**](https://rlbarter.github.io/superheat/) and for examples see the accompanying [**paper by Barter and Yu (2017)**](https://arxiv.org/abs/1512.01524). 


You can install the latest version with:

```{r eval = FALSE}
install.packages("superheat")
```

Stay tuned for new versions with added features and minor usability tweaks.

## Example usage

Usage is straightforward and intuitive:

```{r, fig.align='center', fig.height = 7, fig.width = 9}
library(superheat)
set.seed(1347983)
selected.rows <- sample(1:nrow(mtcars), 10)
# add a space after the column names for aesthetic purposes
colnames(mtcars) <- paste0(colnames(mtcars), " ")

# plot the superheatmap
superheat(mtcars[selected.rows,], 
          # scale columns
          scale = T, 
          # label aesthetics
          left.label.size = 0.5,
          bottom.label.size = 0.15,
          bottom.label.text.angle = 90,
          bottom.label.text.alignment = "right",
          bottom.label.col = "white",
          # dendrogram
          row.dendrogram = T,
          # top plot
          yt = sapply(mtcars, function(x) cor(x, mtcars$hp)),
          yt.plot.type = "bar",
          yt.axis.name = "correlation\nwith\nhorsepower",
          # column order
          order.cols = order(sapply(mtcars, function(x) cor(x, mtcars$hp))),
          # grid lines
          grid.vline.col = "white",
          grid.hline.col = "white")

```