---
title:  "Getting fancy with ggplot2: code for alternatives to grouped bar charts"
author: "Rebecca Barter"
categories: [R, Visualization]
date: 2018-05-29T23:13:14-05:00
type: "post" 
description: "In this post I present the code I wrote to prodocue the figures in my previous post about alternatives to grouped bar charts." 
---


Here I provide the code I used to create the figures from my [previous post on alternatives to grouped bar charts](../2018-05-29_alternatives_dodged_bars/index.html). You are encouraged to play with them yourself!


# Bars 

First, I load in the libraries (the data comes from the `dslabs` library), and convert the data to long-form for creating the grouped bar charts.

```{r message=FALSE, warning=FALSE}
# load in libraries
library(tidyverse)
library(dslabs)
# combine male and female success rates into single data frame
success_rates <- rbind(
  # male success rates
  transmute(research_funding_rates, 
            discipline, 
            success = success_rates_men, 
            gender = "Male"),
  # female success rates
  transmute(research_funding_rates, 
            discipline, 
            success = success_rates_women, 
            gender = "Female")) 
```

```{r}
# print the data
success_rates
```



I used `geom_bar()` to create a grouped bar chart containing the data. I grouped by gender by setting `fill = gender` within the `aes()` function.

```{r message=FALSE, warning=FALSE, fig.align="center"}
# make grouped bar plot
ggplot(success_rates) +
  # add bar for each discipline colored by gender
  geom_bar(aes(x = discipline, y = success, fill = gender),
           stat = "identity", position = "dodge") +
  # name axes and remove gap between bars and y-axis
  scale_y_continuous("Success Rate", expand = c(0, 0)) +
  scale_x_discrete("Discipline") +
  scale_fill_manual(values = c("#468189", "#9DBEBB")) +
  # remove grey theme
  theme_classic(base_size = 18) +
  # rotate x-axis and remove superfluous axis elements
  theme(axis.text.x = element_text(angle = 90, 
                                   hjust = 1, vjust = 0),
        axis.line = element_blank(),
        axis.ticks.x = element_blank()) 
```

# Slope plot

To create the sloped chart, it ends up being easier to keep the data in its original wide form.


The first element I add is the slopes from men to women using `geom_segment()`. Each slope is colored based on whether men or women are more successful.

```{r message=FALSE, warning=FALSE, fig.height = 8, fig.align="center"}
gg_slope <- research_funding_rates %>%
  # add a variable for when men are more successful than women (for colours)
  mutate(men_more_successful = success_rates_men > success_rates_women) %>%
  ggplot() +
  # add a line segment that goes from men to women for each discipline
  geom_segment(aes(x = 1, xend = 2, 
                   y = success_rates_men, 
                   yend = success_rates_women,
                   group = discipline,
                   col = men_more_successful), 
               size = 1.2) +
  # set the colors
  scale_color_manual(values = c("#468189", "#9DBEBB"), guide = "none")  +
  # remove all axis stuff
  theme_classic() + 
  theme(axis.line = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank()) 
gg_slope
```

Next, I add a left and right axis at the start and end of the slopes.

```{r message=FALSE, warning=FALSE, fig.height = 8, fig.align="center"}
gg_slope <- gg_slope +
  # add vertical lines that act as axis for men
  geom_segment(x = 1, xend = 1, 
               y = min(research_funding_rates$success_rates_men) - 2,
               yend = max(research_funding_rates$success_rates_men) + 1,
               col = "grey70", size = 0.5) +
  # add vertical lines that act as axis for women
  geom_segment(x = 2, xend = 2, 
               y = min(research_funding_rates$success_rates_men) - 2,
               yend = max(research_funding_rates$success_rates_men) + 1,
               col = "grey70", size = 0.5) +
  # add the words "men" and "women" above their axes
  geom_text(aes(x = x, y = y, label = label),
            data = data.frame(x = 1:2, 
                              y = 2 + max(research_funding_rates$success_rates_men),
                              label = c("men", "women")),
            col = "grey30",
            size = 6) 
gg_slope  
```

I also need to add an identifier for each discipline's slope. I do this by annotating the plot with text.

```{r message=FALSE, warning=FALSE, fig.height = 8, fig.align="center"}
gg_slope <- gg_slope +
  # add the label and success rate for each discipline next the men axis
  geom_text(aes(x = 1 - 0.03, 
                y = success_rates_men, 
                label = paste0(discipline, ", ", 
                               round(success_rates_men, 1), "%")),
             col = "grey30", hjust = "right") +
  # add the success rate next to each point on the women axis
  geom_text(aes(x = 2 + 0.08, 
                y = success_rates_women, 
                label = paste0(round(success_rates_women, 1), "%")),
            col = "grey30") +
  # set the limits of the x-axis so that the labels are not cut off
  scale_x_continuous(limits = c(0.5, 2.1)) 
gg_slope
```

Finally, I decide to add points (for aesthetic purposes). Behind each circle, I add a slightly larger white circle to act as a border and to give a slight "gap" look.

```{r message=FALSE, warning=FALSE, fig.height = 8, fig.align="center"}
gg_slope <- gg_slope + 
  
  # add the white outline for the points at each rate for men
  geom_point(aes(x = 1, 
                 y = success_rates_men), size = 4.5,
             col = "white") +
  # add the white outline for the points at each rate for women
  geom_point(aes(x = 2, 
                 y = success_rates_women), size = 4.5,
             col = "white") +
  
  # add the actual points at each rate for men
  geom_point(aes(x = 1, 
                 y = success_rates_men), size = 4,
             col = "grey60") +
  # add the actual points at each rate for men
  geom_point(aes(x = 2, 
                 y = success_rates_women), size = 4,
             col = "grey60") 
gg_slope
```

# Horizontal dots

To create the horizontal dot plot, I again keep the data in its original wide-form but I arrange the disciplines in order of women's success rate (this will make the plot easier to read).

The first things I do to create the dot plot is add the horizontal discipline lines and the points for the success rates of men and women. 


```{r message = FALSE, warning = FALSE, fig.width = 8, fig.align="center"}
gg_dot <- research_funding_rates %>%
  # rearrange the factor levels for discipline by rates for women
  arrange(success_rates_women) %>%
  mutate(discipline = fct_inorder(discipline)) %>%
  
  ggplot() +
  # remove axes and superfluous grids
  theme_classic() +
  theme(axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line = element_blank()) +
  
  
  # add a dummy point for scaling purposes
  geom_point(aes(x = 12, y = discipline), 
             size = 0, col = "white") + 
  
  # add the horizontal discipline lines
  geom_hline(yintercept = 1:9, col = "grey80") +
  
  # add a point for each male success rate
  geom_point(aes(x = success_rates_men, y = discipline), 
             size = 11, col = "#9DBEBB") +
  # add a point for each female success rate
  geom_point(aes(x = success_rates_women, y = discipline),
             size = 11, col = "#468189") 
gg_dot

```

Since I really like to annotate my figures with text so that the audience can work less hard, I add the success rate on top of each dot. I also prefer to annotate features my plot directly instead of using legends, so I also add a label for men and women.

```{r message = FALSE, warning = FALSE, fig.width = 8, fig.align="center"}
gg_dot <- gg_dot + 
  # add the text (%) for each male success rate
  geom_text(aes(x = success_rates_men, y = discipline, 
                label = paste0(round(success_rates_men, 1))),
            col = "black") +
  # add the text (%) for each female success rate
  geom_text(aes(x = success_rates_women, y = discipline, 
                label = paste0(round(success_rates_women, 1))),
            col = "white") +
  # add a label above the first two points
  geom_text(aes(x = x, y = y, label = label, col = label),
            data.frame(x = c(25.6 - 1.1, 26.5 + 0.6), y = 10, 
                       label = c("women", "men")), size = 6) +
  scale_color_manual(values = c("#9DBEBB", "#468189"), guide = "none") +
  
  # manually specify the x-axis
  scale_x_continuous(breaks = c(15, 20, 25), 
                     labels = c("15%", "20%", "25%")) +
  # manually set the spacing above and below the plot
  scale_y_discrete(expand = c(0.2, 0)) 
gg_dot
```


# References

- The research funding data comes from Rafael Irizarry's `dslabs` R package

- The slope and dot plot ideas for this post come from [Ann K. Emery's post](http://annkemery.com/clustered-bar-chart/) on the same topic. 