---
title: "Interactive Visualization in R"
author: "Rebecca Barter"
date: "4/7/2017"
output: ioslides_presentation
---

# Why be interactive?



## Examples on the web

Some super cool examples include:

- [Olympic medals](https://www.washingtonpost.com/graphics/sports/olympics/the-1000-medals-of-the-united-states/)

- [Vaccination simulation](https://www.theguardian.com/society/ng-interactive/2015/feb/05/-sp-watch-how-measles-outbreak-spreads-when-kids-get-vaccinated)

- [Income](https://flowingdata.com/2016/06/28/distributions-of-annual-income/)

- [America's opinions](https://projects.fivethirtyeight.com/obama-polling-trends/)

- [Music](http://www.datasketch.es/december/code/nadieh/)

- [Sam talks too much](http://www.datasketch.es/july/code/nadieh/)





## Main tools in R

Make your scatter plots, line plots, bar plots, etc interactive

- Plotly

- Highcharts

- crosstalk

Make D3.js plots in R (no javascript required!)

- rCharts (D3.js)

- d3scatter

- networkD3

# Making ggplot2 interactive

## ggplot2

```{r, message=FALSE, warning = FALSE, fig.align="center", fig.height = 4, fig.width = 7}
library(ggplot2)
g <- ggplot(txhousing, aes(x = date, y = sales, group = city)) +
  geom_line(alpha = 0.4)
g
```


## ggplot2 + plotly
```{r, eval = FALSE, message=FALSE, warning = FALSE, fig.align="center", fig.height = 4, fig.width = 7}
library(plotly)
g <- ggplot(txhousing, aes(x = date, y = sales, group = city)) +
  geom_line(alpha = 0.4) 
ggplotly(g, tooltip = c("city"))
```

<center>
```{r, echo = FALSE, message=FALSE, warning = FALSE, fig.align="center", fig.height = 4, fig.width = 7}
library(plotly)
g <- ggplot(txhousing, aes(x = date, y = sales, group = city)) +
  geom_line(alpha = 0.4) 
ggplotly(g, tooltip = c("city"))
```
</center>




## plotly



```{r eval = FALSE}
g <- txhousing %>% 
  # group by city
  group_by(city) %>%
  # initiate a plotly object with date on x and median on y
  plot_ly(x = ~date, y = ~median) %>%
  # add a line plot for all texan cities
  add_lines(name = "Texan Cities", hoverinfo = "none", 
            type = "scatter", mode = "lines", 
            line = list(color = 'rgba(192,192,192,0.4)')) %>%
  # plot separate lines for Dallas and Houston
  add_lines(name = "Houston", 
            data = filter(txhousing, 
                          city %in% c("Dallas", "Houston")),
            hoverinfo = "city",
            line = list(color = c("red", "blue")),
            color = ~city)
g
```

## plotly


<center>
```{r echo = FALSE, warning=FALSE, fig.height = 5, fig.width = 8}
g <- txhousing %>% 
  # group by city
  group_by(city) %>%
  
  # initiate a plotly object with date on x and median on y
  plot_ly(x = ~date, y = ~median) %>%
  
  add_lines(name = "Texan Cities", hoverinfo = "none", 
          type = "scatter", mode = "lines", 
          line = list(color = 'rgba(192,192,192,0.3)')) %>%
  # plots one line per city since p knows city is a grouping variable
  add_lines(name = "Houston", 
            data = filter(txhousing, city %in% c("Dallas", "Houston")),
            color = ~city, hoverinfo = "city",
            line = list(color = c("red", "blue")))
g
```
</center>

## plotly: adding a slider

```{r eval = FALSE, warning=FALSE}
g %>% rangeslider
```

<center>
```{r echo = FALSE, fig.height = 5, fig.width = 8, warning=FALSE}
g %>% rangeslider()
```
</center>


# Linking with Crosstalk


## Linking with Crosstalk



```{r eval = FALSE}
library(crosstalk)
# define a shared data object
d <- SharedData$new(mtcars)
# make a scatterplot of disp vs mpg
scatterplot <- plot_ly(d, x = ~mpg, y = ~disp) %>%
  add_markers(color = I("navy"))
# define two subplots: boxplot and scatterplot
subplot(
  # boxplot of disp
  plot_ly(d, y = ~disp) %>% 
    add_boxplot(name = "overall", 
                color = I("navy")),
  # scatterplot of disp vs mpg
  scatterplot, 
  shareY = TRUE, titleX = T) %>% 
  layout(dragmode = "select")


```



## Linking with Crosstalk

```{r echo = FALSE}
library(crosstalk)
# define a shared data object
d <- SharedData$new(mtcars)
# make a scatterplot of disp vs mpg
scatterplot <- plot_ly(d, x = ~mpg, y = ~disp) %>%
  add_markers(color = I("navy"))
# define two subplots: boxplot and scatterplot
subplot(
  # boxplot of disp
  plot_ly(d, y = ~disp) %>% 
    add_boxplot(name = "overall", 
                color = I("navy")),
  # scatterplot of disp vs mpg
  scatterplot, shareY = TRUE, titleX = T) %>% 
  layout(dragmode = "select")

```

## Linking with Crosstalk

```{r, message=FALSE, eval = FALSE}
# make subplots
p <- subplot(
  # histogram (counts) of gear
  plot_ly(d, x = ~factor(gear)) %>% 
    add_histogram(color = I("grey50")),
  # scatterplot of disp vs mpg
  scatterplot, 
  titleX = T
) 
layout(p, barmode = "overlay")
```



## Linking with Crosstalk

```{r, message=FALSE, echo = FALSE}
p <- subplot(
  plot_ly(d, x = ~factor(gear)) %>% 
    add_histogram(color = I("grey50")),
  scatterplot, titleX = T
) 
layout(p, barmode = "overlay")
```

# Easy D3.js in R

## Force networks

```{r}
library(networkD3)
data(MisLinks, MisNodes)
head(MisLinks, 3)
head(MisNodes, 3)
```

## Force networks

```{r eval = FALSE}
forceNetwork(Links = MisLinks, Nodes = MisNodes, Source = "source",
             Target = "target", Value = "value", NodeID = "name",
             Group = "group", opacity = 0.9, Nodesize = 3, 
             linkDistance = 100, fontSize = 20)
```

## Force networks

```{r, echo = FALSE}
library(networkD3)
data(MisLinks, MisNodes)
forceNetwork(Links = MisLinks, 
             Nodes = MisNodes, 
             Source = "source", Target = "target", 
             Value = "value", NodeID = "name",
             Group = "group", opacity = 0.9, Nodesize = 3, 
             linkDistance = 100, fontSize = 20)
```

## Chord diagrams

```{r echo = FALSE}
hair.preference <- matrix(c(11975,  5871, 8916, 2868,
              1951, 10048, 2060, 6171,
              8010, 16145, 8090, 8045,
              1013,   990,  940, 6907),
            byrow = TRUE, nrow = 4, ncol = 4)
haircolors <- c("black", "blonde", "brown", "red")
dimnames(hair.preference) <- list(have = haircolors,
                    prefer = haircolors)

```

```{r eval = FALSE}
# devtools::install_github("mattflor/chorddiag")
library(chorddiag)
# plot the hair preferences data
```
```{r echo = FALSE}
library(chorddiag)
```
```{r}
hair.preference
```
```{r eval = FALSE}
# make a chord diagram
groupColors <- c("#000000", "#FFDD89", "#957244", "#F26223")
chorddiag(hair.preference, 
          groupColors = groupColors, groupnamePadding = 40)
```

## Chord diagrams

```{r echo = FALSE}
groupColors <- c("#000000", "#FFDD89", "#957244", "#F26223")
chorddiag(hair.preference, groupColors = groupColors, 
          groupnamePadding = 40)

```

## References

plotly references:

https://cpsievert.github.io/plotly_book/

https://plot.ly/r/

crosstalk:

https://rstudio.github.io/crosstalk/using.html

chord diagram example: 

http://stackoverflow.com/questions/14599150/chord-diagram-in-r
