---
title:  "Alternatives to grouped bar charts"
author: "Rebecca Barter"
categories: [R, Visualization]
date: 2018-05-29T21:13:14-05:00
type: "post" 
description: "One of the most common chart types that is simultaneously the most difficult to read is the grouped bar chart. Fortunately, there exist several substantially more effective alternatives that convey the same information without overwhelming our visual cognition abilities." 
---


At some point in your life you have probably found yourself standing face-to-face with a beast known as a grouped bar chart. Perhaps it was in a research paper where the authors were trying to compare the results of several models across different datasets, or it was in a talk given by a colleague who was trying to compare the popularity of different products among distinct groups of consumers. 

The first time you encountered a grouped bar chart you might have thought "what a neat way to put so much information in a single plot!". However, the moment you started trying to see whether the orange bar always does better than the green bar but worse than the blue bar.

In the example below (made with ggplot2), I show a simple grouped bar chart comparing research funding success rates across a variety of disciplines with the data split into two categories for each discipline: males and females. This data came from Rafael Irizarry's excellent R package, `dslabs`, containing datasets for teaching data science. 


```{r echo = FALSE, message=FALSE, warning=FALSE}
# load in libraries
library(tidyverse)
library(dslabs)
library(knitr)
# combine male and female success rates into single data frame
success_rates <- rbind(
  # male success rates
  transmute(research_funding_rates, 
            discipline, 
            success = success_rates_men, 
            gender = "Male"),
  # female success rates
  transmute(research_funding_rates, 
            discipline, 
            success = success_rates_women, 
            gender = "Female"))
```

```{r echo = FALSE, message=FALSE, warning=FALSE, fig.align="center", fig.width = 10}
# make grouped bar plot
ggplot(success_rates) +
  # add bar for each discipline colored by gender
  geom_bar(aes(x = discipline, y = success, fill = gender),
           stat = "identity", position = "dodge") +
  # name axes and remove gap between bars and y-axis
  scale_y_continuous("Success Rate", expand = c(0, 0)) +
  scale_x_discrete("Discipline") +
  scale_fill_manual(values = c("#468189", "#9DBEBB")) +
  # remove grey theme
  theme_classic(base_size = 18) +
  # rotate x-axis and remove superfluous axis elements
  theme(axis.text.x = element_text(angle = 90, 
                                   hjust = 1, vjust = 0),
        axis.line = element_blank(),
        axis.ticks.x = element_blank()) 
```

Even in this simple example, I find it really difficult to compare male and female funding success rates across genders. Are the males having more success than the females overall? It's genuinely hard to tell unless I examine each discipline closely and see that men are doing better in the first two groups, but not the two after that, and then I think they are doing a bit better in the categories after that, but now I'm having a hard time and I really want to stop looking at this plot now...

Fortunately, people such as Ann K. Emery (in [her post on grouped bar charts](http://annkemery.com/clustered-bar-chart/)) have thought about how to represent this information in a better way. Below I will show 2 alternative approaches (slope charts and horizontal dot plots). 

While these are not necessarily "standard" plot types (as in there is no specific `geom_` for them), true freedom in visualization comes with the realization that most plots are just arrangements of circles and lines in space. When you truly embrace this idea, you can make any type of figure you want! By means of demonstration, in my [next post I show the ggplot2 R code for making each plot](../2018-05-29_getting_fancy_ggplot2/index.html).

# Slope plots

The first alternative is similar to parallel coordinates plots for multi-variate data visualization. The slope plot contains two axes (or three or four, depending on how many groups you have), one for each group (male and female). The disciplines are each represented by a line connecting the male success rates to the female success rates in as coded by their y-position. This plot makes it very clear for which disciplines women have greater funding success than men (the upward sloping, darker colored lines) and it also makes it very clear which disciplines have higher funding success rates overall (Physics and Chemical sciences).

```{r echo = FALSE, message=FALSE, warning=FALSE, fig.height = 8, fig.align="center"}
research_funding_rates %>%
  # add a variable for when men are more successful than women (for colours)
  mutate(men_more_successful = success_rates_men > success_rates_women) %>%
  ggplot() +
  # add a line segment that goes from men to women
  geom_segment(aes(x = 1, xend = 2, 
                   y = success_rates_men, 
                   yend = success_rates_women,
                   group = discipline,
                   col = men_more_successful), 
               size = 1.2) +
  # add vertical lines that act as axis for men
  geom_segment(x = 1, xend = 1, 
               y = min(research_funding_rates$success_rates_men) - 2,
               yend = max(research_funding_rates$success_rates_men) + 1,
               col = "grey70", size = 0.5) +
  # add vertical lines that act as axis for women
  geom_segment(x = 2, xend = 2, 
               y = min(research_funding_rates$success_rates_men) - 2,
               yend = max(research_funding_rates$success_rates_men) + 1,
               col = "grey70", size = 0.5) +
  # add the white outline for the points at each rate for men
  geom_point(aes(x = 1, 
                 y = success_rates_men), size = 4.5,
             col = "white") +
  # add the white outline for the points at each rate for women
  geom_point(aes(x = 2, 
                 y = success_rates_women), size = 4.5,
             col = "white") +
  # add the actual points at each rate for men
  geom_point(aes(x = 1, 
                 y = success_rates_men), size = 4,
             col = "grey60") +
  # add the actual points at each rate for men
  geom_point(aes(x = 2, 
                 y = success_rates_women), size = 4,
             col = "grey60") +
  # add the label for each discipline next the men axis
  geom_text(aes(x = 1 - 0.03, 
                 y = success_rates_men, 
                label = paste0(discipline, ", ", round(success_rates_men, 1), "%")),
             col = "grey30", hjust = "right") +
  # add the label next to each point on the women axis
  geom_text(aes(x = 2 + 0.08, 
                y = success_rates_women, 
                label = paste0(round(success_rates_women, 1), "%")),
            col = "grey30") +
  # add the words "men" and "women" above their axes
  geom_text(aes(x = x, y = y, label = label),
            data = data.frame(x = 1:2, 
                              y = 2 + max(research_funding_rates$success_rates_men),
                              label = c("men", "women")),
            col = "grey30",
            size = 6) +
  # set the limits of the x-axis
  scale_x_continuous(limits = c(0.5, 2.1)) +
  # set the colors
  scale_color_manual(values = c("#468189", "#9DBEBB"), guide = "none") +
  # universally increase the size of the text and remove grey background
  theme_classic(base_size = 16) + 
  # remove all axis stuff
  theme(axis.line = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank())
```


What we see much more clearly is that the number of disciplines in which men are more successful is similar to the number of disciplines in which women are more successful. The discipline with the most similar success rates between men and women is the Chemical sciences. All of this information was very difficult to obtain from the grouped bar chart!


# Horizontal dot plots

Another alternative is the horizontal dot plot: In this case each discipline is again represented by a line, but instead of coding the success rates as the y-position at the end of the discipline slope, the success rate is coded as the x-position along the discipline-specific horizontal line. I have ordered the disciplines by the women's funding success rate.

```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 8, fig.align="center"}
research_funding_rates %>%
  # rearrange the factor levels for discipline by rates for women
  arrange(success_rates_women) %>%
  mutate(discipline = fct_inorder(discipline)) %>%
  ggplot() +
  # add a dummy point for scaling purposes
  geom_point(aes(x = 12, y = discipline), 
             size = 0, col = "white") + 
  # add the horizontal grid lines
  geom_hline(yintercept = 1:9, col = "grey80") +
  # add a point for each male success rate
  geom_point(aes(x = success_rates_men, y = discipline), 
             size = 11, col = "#9DBEBB") +
  # add the text (%) for each male success rate
  geom_text(aes(x = success_rates_men, y = discipline, 
                label = paste0(round(success_rates_men, 1))),
            col = "black") +
  # add the border for each point for female success rate
  geom_point(aes(x = success_rates_women, y = discipline),
             size = 11.5, col = "white") +
  # add a point for each female success rate
  geom_point(aes(x = success_rates_women, y = discipline),
             size = 11, col = "#468189") +
  # add the text (%) for each female success rate
  geom_text(aes(x = success_rates_women, y = discipline, 
                label = paste0(round(success_rates_women, 1))),
            col = "white") +
  # add a label above the first two points
  geom_text(aes(x = x, y = y, label = label, col = label),
            data.frame(x = c(25.6 - 1.1, 26.5 + 0.6), y = 10, 
                       label = c("women", "men")), size = 6) +
  scale_color_manual(values = c("#9DBEBB", "#468189"), guide = "none") +
  # universally increase size of text and remove grey theme
  theme_classic(base_size = 14) +
  # manually specify the x-axis
  scale_x_continuous(breaks = c(15, 20, 25), 
                     labels = c("15%", "20%", "25%")) +
  # remove gaps in the y-axis
  scale_y_discrete(expand = c(0.2, 0)) +
  # remove axes
  theme(axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line = element_blank()) 
```

For me, the horizontal dot plot highlights that in the disciplines with lower funding rates overall (the bottom few rows), men have higher success rates.

Regardless of your preference, it is fairly clear that both the slope plot and the horizontal dot plots are easier to digest than the grouped bar plot!


# Code

My [next post](../2018-05-29_getting_fancy_ggplot2/index.html) shows the ggplot2 code that I wrote to produce the three plots in this post.


# References

- The research funding data comes from Rafael Irizarry's `dslabs` R package

- The slope and dot plot ideas for this post come from [Ann K. Emery's post](http://annkemery.com/clustered-bar-chart/) on the same topic. 