---
title: "10 R tips from rstudio::conf(2020) - tidy evaluation, piping into later arguments, changing axis colors and more."
author: "Rebecca Barter"
output:
  blogdown::html_page:
    toc: true
categories: [R, rstudioconf, tidyverse, ggplot2, tidyeval, visualization]
date: 2020-02-03T23:13:14-05:00
type: "post" 
description: "This week I had the pleasure of attending rstudio::conf(2020) in San Francisco. Throughout the course of the week I met many wonderful people and learnt many things. This post covers some of the smaller tips and tricks that I learnt throughout the conference." 
---


Rstudio::conf(2020) is my favourite conference. This was my second year attending as a diversity scholar, and I was yet again blown away by the friendliness of the community and the quality of the presentations. Over the course of the week, I met so many wonderful and talented humans, and learnt so many things. While there were number of big things I learnt, this post is all about those little tips and tricks that I picked up from watching everyone's fantastic presentations, and having many insightful conversations. 


# Tip 1: `geom_col()`: you'll never have to specify "stat = identity" for your bar plots ever again!

Most of the time when I want to make a bar chart, I want to provide an `x` variable for the bar identifier, and a `y` variable for the height of the bar. However, when I use the `geom_bar()` function for this, I get the annoying error `Error: stat_count() must not be used with a y aesthetic.`. This happens *literally every time*. The typical fix for this is to include `stat = 'identity'` as an argument of `geom_bar()`, but it turns out *there's a better way*!

During his fantastic talk on his ggtext package, Klaus Wilkinson casually used the `geom_col()` function to create a *column chart* (is that what `geom_col()` stands for?), which, as it turns out, is what I've been trying to make the whole time. To demonstrate `geom_col()` we will use the inbuilt islands dataset consisting of the sizes of the land masses. `islands` initially comes as a named vector, so here are some bonus tips: the `enframe()` function converts a named vector to a two-column data frame, and the `gt()` function is a nice way to print a data frame in html format.

```{r}
library(tidyverse)
library(gt)
large_islands <- islands %>%
  # convert the vector to a data frame
  enframe(name = "island", value = "land_mass") %>%
  # filter to the larger islands
  filter(land_mass > 1000)  
gt(large_islands)
```


When I want to create a bar plot I usually try 

```{r, eval = FALSE}
large_islands %>%
  ggplot() +
  geom_bar(aes(x = island, y = land_mass))
```

and fix the error by specifying `stat = "identity"`. 

```{r, eval = FALSE, fig.align="center"}
large_islands %>%
  ggplot() +
  geom_bar(aes(x = island, y = land_mass), stat = "identity")
```

While it seems like a very small life improvement, you have no idea how much joy simply being able to use `geom_col()` brings me. I'm never going to have to specify `stat = "identity"` again! I'm so pathetically excited!

```{r, eval = FALSE, fig.align="center"}
large_islands %>%
  ggplot() +
  geom_col(aes(x = island, y = land_mass))
```


# Tip 2: Pipe into later arguments of a function using `.`

While I didn't technically learn this one from a talk at rstudio::conf(2020), I did run into an old friend Garth ([@garthtarr](https://twitter.com/garthtarr)) who told me about this little trick: if you want to pipe an object into any argument of a function other than the first one, you can do so using the `.` placeholder. For instance, let's say that you want to use the `lm` function to fit a linear model, and because you, like me, are a tidyverse fangirl, you want to pipe the data into `lm`. 

Using the `midwest` dataset from the ggplo2 package, we will try to model population density (`popdensity`) as a linear function of the poverty rate (`percbelowpoverty`). The following code, however, doesn't work:

```{r, eval = FALSE}
midwest %>% lm(popdensity ~ percbelowpoverty)
```


What's going wrong? There are two things you need to understand about what's happening here:

1. The argument order of `lm()` us `lm(formula, data)`

1. The pipe, `%>%`, automatically places the *object* to the left of the pipe into the first argument of the *function* to the right of the pipe

so `iris %>% lm(Sepal.Length ~ Sepal.Width)` above is equivalent to `lm(formula = iris, data = Sepal.Length ~ Sepal.Width)`, which has the arguments reversed.

Instead of abandoning our trusty pipe (the thought of which fills me with sadness), we can instead pipe the data into the *second* argument using `.`: 

```{r}
midwest %>% lm(popdensity ~ percbelowpoverty, .)
```


which is equivalent to

```{r}
midwest %>% lm(formula = popdensity ~ percbelowpoverty, data = .)
```


Note that we could alternatively just name the formula argument, which would automatically pipe the object into the first *unassigned* argument (which in our case is data), but this becomes cumbersome when you have many arguments. 

```{r}
midwest %>% lm(formula = popdensity ~ percbelowpoverty)
```




# Tip 3: Make gifs using the gifski package





# Tip 4: Usethis and chatty functions


https://sharla.party/post/usethis-for-reporting/


```{r}
library(usethis)
ui_info("hi!")
analysis_year <- 2019
ui_todo("Querying demographics data for
        {analysis_year}...")
```


# Tip 5: Tidy evaluation: accessing variables inside functions using {{ }} and .data

Suppose that you want to write a function that takes in unquoted variable names just like tidyverse functions to.

```{r eval = FALSE}

plotMidwest <- function(var1, var2) {
  ggplot(midwest) +
    geom_point(aes(x = var1, y = var2))
}
plotMidwest(popdensity, percbelowpoverty)
plotMidwest("popdensity", "percbelowpoverty")

plotMidwest <- function(var1, var2) {
  ggplot(midwest) +
    geom_point(aes(x = .data[[var1]], y = .data[[var2]]))
}
plotMidwest(popdensity, percbelowpoverty)
plotMidwest("popdensity", "percbelowpoverty") # works


plotMidwest <- function(var1, var2) {
  ggplot(midwest) +
    geom_point(aes(x = {{ var1 }}, y = {{ var2 }}))
}
plotMidwest(popdensity, percbelowpoverty) # works
plotMidwest("popdensity", "percbelowpoverty")

```





# Tip 6: Adding `NULL` to a plot changes nothing


```{r}
midwest %>%
  ggplot() +
  geom_point(aes(x = popdensity, y = percbelowpoverty)) + 
  NULL
```

Why is this usefull, you ask?

If you ever want to write a function that does something to a plot conditionally based on some feature, it has the potential to save a lot of duplicated code. 

```{r}
plotMidwest <- function(classic = TRUE) {
  if (classic) {
    theme <- theme_classic()
  } else {
    theme <- NULL
  }
midwest %>%
  ggplot() +
  geom_point(aes(x = popdensity, y = percbelowpoverty)) + 
  theme
}

plotMidwest(classic = TRUE)
plotMidwest(classic = FALSE)
```

# Tip 7: Using italics, colour in axes text using the `ggtext` package

More info on ggtext can be found here https://github.com/wilkelab/ggtext



# Tip 9: Using `show_col()` for viewing colour palettes

The scales package does a lot under the hood of ggplot. 

```{r}
library(scales)
show_col(c("#9C89B8", "#F0A6CA", "#EFC3E6", "#F0E6EF"))
```

Also a quick plug for how I chose these colours (and how I choose colors in general): https://coolors.co/app. 

# Tip 10: Using slider for taking rolling averages



